---
title: "Botany 575 Final Project: Suicide Rates, 1987-2016"
author: "Aryan Adhlakha, Kashin Shah, Sai Bulusu"
date: "4/22/2020"
output: html_document
---

# **Introduction and background:**

Suicides are tragedies that leave everlasting effects. They affect families and communities. According to the World Health Organisation, there are close to 800,000 people who commit suicide every year[1]. It is an alarming, but preventable problem. With evidence for trends, we can find the more vulnerable groups and adapt targeted preventive measures to lower the rate. While the link has been higher between mental disorders and suicides historically, there are several other factors that play a role in setting these trends. The central aim is analyse who is more at risk, and under which categories the rates peak, as we compare and analyse the socio-economic information with suicide rates by year and country. 

This dataset, adapted from kaggle, showcases information of suicide rates by country, gdp per capita, age, sex, and generation. With this dataset, we have broken down our questions to analyse each factor of suicide rate and find any correlations. 

>The questions we aim to analyse are:

1. Does age affect the suicide rate?
2. Does sex affect the suicide rate?
3. Correlation between GDP and population suicide rates? A follow-up question to that is, does GDP strongly relate to a certain age-group committing suicides?
4. What is the role of generation in the rates we are examining?
5. Which country has the highest suicide rate, per age group, per generation, and through the years? Can it be related to any factor in specific. Answers may be present in the data or may require additional background research.
6. What percentage of the population is affected by it? Was the rate extremely high or low during a period of time? Examine. Does confounding possibly appear? Explain.

# **Data description:**
This compiled dataset pulled from four other datasets linked by time and place, and was built to find signals correlated to increased suicide rates among different cohorts globally, across the socio-economic spectrum. 
The dataset can be found on https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016. Other datasets that have been used in this accumulating the data are mentioned at the end.
The dataset consists of columns that contain information about the country, year, sex, age group, count of suicides (suicides_no), population, suicide rate (suicide/100k population), country-year composite key, HDI for year, gdp_for_year, gdp_per_capita, generation (based on age grouping average), and are labelled as mentioned.
Using these values, we wish to answer questions regarding any trends we may observe regarding suicide rates. 



```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

## R Markdown

```{r read-transform-data, include = TRUE, echo=FALSE}
data = read_csv("master 2.csv")
```
### 1. Does age affect the suicide rate?

```{r, echo=FALSE}
ageAndSuicideRate = data %>% 
  select (country, year, age, `suicides/100k pop`) %>%
  group_by(age, year) %>%
  summarize(TotalSuicideRate=sum(`suicides/100k pop`))

ggplot(data = ageAndSuicideRate, aes(x = `year`, y = TotalSuicideRate, color=age)) + geom_point() + facet_wrap(~age) + geom_smooth(method = "lm", color='black') + ylab("Suicides per 100 thousand people") + xlab("Through the years") + ggtitle("Suicide numbers vs Years, for different age groups")

totalDeathsPerAge = ageAndSuicideRate %>%
  select(age, TotalSuicideRate) %>%
  group_by(age) %>%  
  summarize(TotalDeaths=sum(TotalSuicideRate)) %>%
  mutate(Percentage=paste(as.character(round(TotalDeaths/sum(TotalDeaths)*100)) ,"%"))

ggplot(totalDeathsPerAge, aes(x=age, y=Percentage)) +
  geom_bar(stat="identity", width=1, color="white", fill = "light blue") +
  ggtitle("Proportion of deaths/age group since 1987 to 2016") + geom_text(aes(label=Percentage), position=position_dodge(width=0.9), vjust=-0.25)

totalDeathsPerAge = ageAndSuicideRate %>%
  select(age, TotalSuicideRate) %>%
  group_by(age) %>%  
  summarize(TotalDeaths = sum(TotalSuicideRate)) %>%
  mutate(Percentage = paste(as.character(round(TotalDeaths/sum(TotalDeaths)*100)) ,"%")) %>% 
  mutate(Affected = TotalDeaths/sum(TotalDeaths)*100) %>%
  mutate(Unaffected = 100 - Affected) %>%
  select(Affected, Unaffected)

chisq.test(totalDeathsPerAge)
```
Through observation, one can notice that people aged 75 years or older have the highest suicide rate and the rate decreases as we go down the age groups. For the chi-square test, we tested whether age and suicide rates are independent. Assuming alpha = 0.05, with a p-value = 5.719e-07 we cannot reject the null and conclude that there is a correlation age and suicide rate.

### 2. Does sex affect the suicide rate?
```{r, echo=FALSE}
sexAndSuicideRate = data %>%
  select(sex, `suicides/100k pop`, year) %>%
  group_by(sex, year) %>%
  summarize(Mean = mean(`suicides/100k pop`)) %>%
  pivot_wider(names_from = "sex", values_from = "Mean") %>%
  mutate(difference = male - female)

ggplot(data = sexAndSuicideRate) + 
    geom_point(mapping = aes(x = factor(year), y = difference), stat = "identity") +
    xlab("Year") + 
    ylab("Difference between male and female rates") + 
    ggtitle("Suicide Rate vs Sex") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

sexAndSuicideRate = sexAndSuicideRate %>%
  select(male, female)

```

### 3. Correlation between GDP and population suicide rates? A follow-up question to that is, does GDP strongly relate to a certain age-group committing suicides?
```{r GDP to population suicide rate,  echo=FALSE}
gdpToSuicideRate = data %>%
  select(`suicides/100k pop`, `gdp_per_capita ($)`, age)

ggplot(data = gdpToSuicideRate, aes(x = `gdp_per_capita ($)`, y = `suicides/100k pop`)) + geom_point()  + ylab("Suicides per 100 thousand people") + xlab("GDB per capita") + ggtitle("Suicide counts vs GDP per capita")

ggplot(data = gdpToSuicideRate, aes(x = `gdp_per_capita ($)`, y = `suicides/100k pop`)) + geom_point() + facet_wrap(~age)  + ylab("Suicides per 100 thousand people") + xlab("GDB per capita") + ggtitle("Suicide counts vs GDP per capita separated by age")

gdpToSuicideRate = gdpToSuicideRate %>%
  select(`suicides/100k pop`)

chisq.test(gdpToSuicideRate)
```
While it appears that there are more cases of suicide when the gdp per capita is low, there is actually no effective correlation between the 2 variables. In terms of age, it seems that the suicide rate increases with age. For the chi-squared test, assuming alpha = 0.05, with a p-value < 2.2 * 10^-16 < alpha, we cannot reject the null and conclude that there is a correlation between gdp and suicide rate.
 
### 4. What is the role of generation in the rates we are examining?
```{r,  echo=FALSE}
generationAndSuicideRate = data %>% 
  select (country, year, generation, `suicides/100k pop`) %>%
  group_by(generation, year) %>%
  summarize(TotalSuicideRate=sum(`suicides/100k pop`))

ggplot(data = generationAndSuicideRate, aes(x = `year`, y = TotalSuicideRate, color=generation)) + geom_point() + facet_wrap(~generation) + geom_smooth(method = "lm", color='black') + ylab("Suicides per 100 thousand people") + xlab("Through the years") + ggtitle("Suicide Rate vs Years, for different generations")

totalDeathsPerGeneration = generationAndSuicideRate %>%
  select(generation, TotalSuicideRate) %>%
  group_by(generation) %>%  
  summarize(TotalDeaths=sum(TotalSuicideRate)) %>%
  mutate(Percentage=paste(as.character(round(TotalDeaths/sum(TotalDeaths)*100)),"%"))

ggplot(totalDeathsPerGeneration, aes(x=generation, y=Percentage)) +
  geom_bar(stat="identity", width=1, color="white", fill = "light PINK") +
  ggtitle("Proportion of deaths/generation since 1987 to 2016")  + geom_text(aes(label=Percentage), position=position_dodge(width=0.9), vjust=-0.25)

totalDeathsPerGeneration = ageAndSuicideRate %>%
  select(age, TotalSuicideRate) %>%
  group_by(age) %>%  
  summarize(TotalDeaths=sum(TotalSuicideRate)) %>%
  mutate(Percentage=paste(as.character(round(TotalDeaths/sum(TotalDeaths)*100)) ,"%")) %>% 
  mutate(Affected=TotalDeaths/sum(TotalDeaths)*100)

totalDeathsPerGeneration = totalDeathsPerGeneration %>%
  mutate(Unaffected = 100 - Affected) %>%
  select(Affected, Unaffected)

chisq.test(totalDeathsPerGeneration)
```
As one can notice, the "Silent" generation was the one that saw most number of suicides across the globe. With a chi-square test, we have a p-value of 5.719 * 10^-07. Assuming alpha = 0.05, we cannot reject the null, and we conclude that there is a correlation between the Suicde rate and generation.

### 5. Which country has the highest suicide rate, per age group, per generation, and through the years? Can it be related to any factor in specific. Answers may be present in the data or may require additional background research.
```{r,  echo=FALSE}
## suicide rate for countries per age group
countryAndAge = data %>%
  select(country, `suicides/100k pop`, age) %>%
  group_by(country, `suicides/100k pop`)


ggplot(countryAndAge, aes(fill = age, y = `suicides/100k pop`, x = country)) + 
    geom_bar(position = "stack", stat = "identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

## suicide rate for countries per generation
countryAndGeneration = data %>%
  select(country, `suicides/100k pop`, generation) %>%
  group_by(country, `suicides/100k pop`)

ggplot(countryAndGeneration, aes(fill = generation, y = `suicides/100k pop`, x = country)) + 
    geom_bar(position = "stack", stat = "identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

## suicide rate for countries year-wise
countryAndYear = data %>%
  select(country, `suicides/100k pop`, year) %>%
  group_by(country, `suicides/100k pop`)


ggplot(countryAndYear, aes(fill = year, y = `suicides/100k pop`, x = country)) + 
    geom_bar(position = "stack", stat = "identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### 6. What percentage of the population is affected by it? Was the rate extremely high or low during a period of time? Examine. Does confounding possibly appear? Explain.

```{r Percentage of the population affected,  echo=FALSE}
populationsByYear = data %>%
  select(year, suicides_no, population) %>%
  group_by(year) %>%
  summarize(affected = sum(suicides_no),
            pop = sum(population)) %>%
  mutate(percent = 100 * affected / pop) %>%
  select(year, percent)

ggplot(data = populationsByYear, aes(x = factor(year), y = percent)) + geom_point() + ylab("Percentage of world population affected") + xlab("Year") + ggtitle("Percentage vs. Year")  + theme(axis.text.x = element_text(angle = 90, hjust = 1))

populationsByYear = populationsByYear %>%
  select(percent)

chisq.test(populationsByYear)
```
Examining the world population and suicide rate, it seems that in 1995, the suicide percentage peaked. We have consistently decreased our rate since then. In our chi-squared test, assuming alpha = 0.05, we have a p-value of 1 > alpha, so we can reject the null and thus conclude that there is no correlation.

# **Conclusion:**
Results thus far match our expectations in some cases, but differ drastically in others. When considering age as a factor, we find that people aged 75 and older peak the rate. The rate then starts to drop as we go further down the age groups. The elderly account for about 12% of the US population, yet we find a peak in this age group. The trend found is that the risk increases with age. With the expectation of suicide rates being higher among males than those in females, the analysis not only proved that, but also showed the vast differences in the numbers. The highest noted percentage difference between males and females is in the year 1995.  We have not found any correlation between gdp per capita and the suicide rate. The generation that peaked the suicide rates was the “Silent” generation (1928 to 1945). This was the generation that was raised during a period of war and economic depression, which can be a strong contributing factor for the rise. 2006 witnessed the most suicides and the rate has since consistently dropped. We also find that amongst the countries, Romania peaked the suicide rate. 

# **References:**

1. who.int/news-room/fact-sheets/detail/suicide
2. United Nations Development Program. (2018). Human development index (HDI). Retrieved from http://hdr.undp.org/en/indicators/137506
3. World Bank. (2018). World development indicators: GDP (current US$) by country:1985 to 2016. Retrieved from http://databank.worldbank.org/data/source/world-development-indicators#
4. [Szamil]. (2017). Suicide in the Twenty-First Century [dataset]. Retrieved from https://www.kaggle.com/szamil/suicide-in-the-twenty-first-century/notebook
5. World Health Organization. (2018). Suicide prevention. Retrieved from http://www.who.int/mental_health/suicide-prevention/en/


