---
title: "script.Rmd"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

## R Markdown

```{r read-transform-data, include = TRUE}
data = read_csv("master 2.csv")
```

3. Correlation between GDP and population suicide rates? A follow-up question to that is, does GDP strongly relate to a certain age-group committing suicides?
```{r GDP to population suicide rate}
gdpToSuicideRate = data %>%
  select(`suicides/100k pop`, `gdp_per_capita ($)`, age)

ggplot(data = gdpToSuicideRate, aes(x = `gdp_per_capita ($)`, y = `suicides/100k pop`)) + geom_point() + geom_smooth(method = "lm") + ylab("Suicides per 100 thousand people") + xlab("GDB per capita") + ggtitle("Suicide counts vs GDP per capita")

ggplot(data = gdpToSuicideRate, aes(x = `gdp_per_capita ($)`, y = `suicides/100k pop`)) + geom_point() + facet_wrap(~age) + geom_smooth(method = "lm") + ylab("Suicides per 100 thousand people") + xlab("GDB per capita") + ggtitle("Suicide counts vs GDP per capita separated by age")
```
While it appears that there are more cases of suicide when the gdp per capita is low, there is actually no effective correlation between the 2 variables. In terms of age, it seems that the suicide rate increases with age.


6. What percentage of the population is affected by it?
Was the rate extremely high or low during a period of time? Examine. 
Does confounding possibly appear? Explain.

```{r Percentage of the population affected}
populationsByYear = data %>%
  select(year, suicides_no, population) %>%
  group_by(year) %>%
  summarize(affected = sum(suicides_no),
            pop = sum(population)) %>%
  mutate(percent = 100 * affected / pop) %>%
  select(year, percent)

ggplot(data = populationsByYear, aes(x = year, y = percent)) + geom_point() + ylab("Percentage of world population affected") + xlab("Year") + ggtitle("Percentage vs. Year")
```
Examining the world population and suicide rate, it seems that in 2006, the suicide percentage peaked. We have consistently decreased our rate since then.
